Abstractions for Type-Level Programming

### 2022-03-19 slides.tex
m
Update slides

\begin{slide}{Example 1: Database queries (2016)}
\begin{center}
\includegraphics[width=0.3\textwidth]{figures/spark.pdf}
\end{center}
\end{slide}

\begin{slide}{Example 1: Database queries (2016)}
Spark APIs are practically untyped:
\begin{code}
class DataFrame {
  /** Inner equi-join with another DataFrame on the given column.
   *  The join column will only appear once in the output. */
  def join(right: DataFrame, column: String): DataFrame
}
\end{code}
\pause
Revisited with type-level programming:
\begin{code}
class DF[X] {
  def join[Y](df: DF[Y], col: String): DF[col+(X-col)++(Y-col)]
}
\end{code}
\end{slide}

\begin{slide}{Example 2: Regular expressions (2022)}
Regular expressions in Scala's standard library:
\begin{code}
val rational = new Regex("(\\d+)\\.?(\\d+)?")
rational.unapply("3.1415"): Option[Seq[String]]
\end{code}
\pause
Revisited with type-level programming:
\begin{code}
rational.unapply("3.1415"): Option[(String, Option[String])]
\end{code}
\pause
\begin{code}
class Regex(pattern: String) {
  def unapply(s: String): Option[GroupsOf[@pattern@]]
}
\end{code}
\end{slide}

@@

\begin{slide}{What changed between 2016 and 2022?}
Example 1 uses a hack, implicits:
\begin{itemize}
  \item convoluted to use
  \item slow to compile
\end{itemize}

Example 2 uses a new language construct, match types, one of the main contributions of this thesis.
\end{slide}

@@

\begin{slide}{Generalized singleton types: A proposal}
Our proposal consists of 3 changes:
\begin{enumerate}
  \item new types for if-then-else, pattern matching, constructors, functions calls
  \item "precise mode" of type inference
  \item type evaluation, during subtyping
\end{enumerate}
\end{slide}

\begin{slide}{New types}
What's the type of | if (x == 0) "zero" else "one"|?
\begin{itemize}
  \pause\item |String| ~(Scala 2)
  \pause\item |"zero"|~{\footnotesize\texttt\textbar}~|"one"| ~(Scala 3)
  \pause\item |If[x.type == 0,|| "zero", "one"]| ~(proposed)
  \pause\item |{ if (x == 0) "zero" else "one" }| ~(proposed, syntactic sugar)
\end{itemize}
\pause
Similarly, we add new types for other constructs: |Match[]|, |New[]| and |Call[]|, and the corresponding syntactic sugar.
\end{slide}

@@

\begin{slide}{"Precise mode" of type inference: list concatenation example}
\begin{code}
dependent def concat(xs: List, ys: List) <: List =
  xs match
    case x :: xs => x :: concat(xs, ys)
    case Nil => ys
\end{code}
\pause
\begin{overprint}
\onslide<1-2>
\begin{code}
dependent val l1 = "A" :: Nil
dependent val l2 = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { "A" :: "B" :: Nil }
\end{code}

@@

More interesting for pattern matching: we "desugar" pattern matching expressions into if-then-else \& type tests.
\pause
\smallskip

We evaluate type tests using subtyping and type disjointness:
\begin{code}
eval(@\tp{x.asInstanceOf}@[Foo]) = true   iff  x.type <: Foo
eval(@\tp{x.asInstanceOf}@[Foo]) = false  iff  disj(x.type, Foo)
\end{code}
\end{slide}

\begin{slide}{Type evaluation: pattern matching example}
\begin{overprint}
\onslide<1>
\begin{code}
dependent val foo(x: Any) =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<2>
\begin{code}
dependent val foo(x: Any): {
  x match
    case s: String => @\tp{s}@
    @\tp{case i: Int}@ => @\tp{i+1}@
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<3->
\begin{code}
dependent val foo(x: Any): {
  if (x.isInstanceOf[String]) x.asInstanceOf[String]
  @\tp{else if }@(@\tp{x.isInstanceOf}@[@\tp{Int}@]) @\tp{x.asInstanceOf}@[Int] @\tp{+ 1}@
  @\tp{else throw new}@ MatchError()
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\end{overprint}
\bigskip

\begin{overprint}
\onslide<4>
|foo(42): { foo(42) }|
\onslide<5>
|foo(42): { 42.asInstanceOf[Int] + 1 }|
\onslide<6>
|foo(42): { 42 + 1 }|
\onslide<7>
|foo(42): { 43 }|
\onslide<8->
|foo(42): 43|
\end{overprint}

\begin{overprint}
\onslide<9>
|foo(readString())|
\onslide<10>
|foo(readString()): { foo(_: String) }|
\onslide<11>
|foo(readString()): { (_: String).asInstanceOf[String] }|
\onslide<12>
|foo(readString()): { (_: String) }|
\onslide<13>
|foo(readString()): String|
\end{overprint}
\end{slide}

\begin{slide}{Generalized Singletons: Recap}

We proposed the following 3 changes:
\begin{enumerate}
  \item lift a subset of Scala's language constructs to the type level
  \item add a "precise mode" of type inference
  \item evaluate those types during subtyping
\end{enumerate}
\end{slide}

TODO
\end{slide}

@@

\begin{slide}{Overview}
\begin{code}
type Elem[X] = X match
  case String => Char
  case List[t] => Elem[t]
  case Any => X
\end{code}
\end{slide}

\begin{slide}{Disjointness}
|Elem[Seq[Int]]| is stuck
\end{slide}

\begin{slide}{Relating match terms and types: a defining property}
|Elem[Seq[Int]]| is stuck
\end{slide}

\begin{slide}{Formalizing Match Types}
System FM is an extension of System \Fsub with
\begin{itemize}
  \item pattern matching
  \item opaque classes
  \item match types
\end{itemize}
Design goals:
\begin{itemize}
  \item explain the essence of match types
  \item give us confidence in our design
  \item be as simple as possible
\end{itemize}
\end{slide}

@@

\begin{tabular}{@{}llll@{}}
|class A|          &$\C = \{\A\}$     & $\Psi = \varnothing$  & $\Xi = \varnothing$\\
|class A; class B extends @\tp{A}@| &$\C = \{\A, \B\}$ & $\Psi = \{(\B, \A)\}$ & $\Xi = \varnothing$\\
|class A; class D| &$\C = \{\A, \D\}$ & $\Psi = \varnothing$  & $\Xi = \{(\A, \D)\}$\\
|class A; trait T| &$\C = \{\A, \T\}$ & $\Psi = \varnothing$  & $\Xi = \varnothing$\\
\end{tabular}
\end{slide}

@@

\begin{slide}{Empty Types}
Our system does not like empty types.

@@

\begin{slide}{Match Types, Conclusion}
TODO
\end{slide}

@@

\begin{slide}{Type-Safe Regular Expressions (simplified)}
\begin{code}
val rational = new Regex("(\\d+)\\.(\\d+)")
rational.unapply("3.1415"): Option[(String, String)]

@@

% \appendix
\end{document}

>>>>

\begin{slide}{Example 1: database queries (2016)}
\begin{center}
\includegraphics[width=0.3\textwidth]{figures/spark.pdf}
\end{center}
\end{slide}

\begin{slide}{Example 1: database queries (2016)}
Spark APIs are practically untyped:
\begin{code}
class DataFrame {
  /** Inner equi-join with another DataFrame on the given column.
   *  The join column will only appear once in the output. */
  def join(right: DataFrame, column: String): DataFrame
}
\end{code}
\pause
Revisited with type-level programming:
\begin{code}
class DF[X] {
  def join[Y](df: DF[Y], col: String): DF[col+(X-col)++(Y-col)]
}
\end{code}
\end{slide}

\begin{slide}{Example 2: regular expressions (2022)}
Regular expressions in Scala's standard library:
\begin{code}
val rational = new Regex("(\\d+)\\.?(\\d+)?")
rational.unapply("3.1415"): Option[Seq[String]]
\end{code}
\pause
Revisited with type-level programming:
\begin{code}
rational.unapply("3.1415"): Option[(String, Option[String])]
\end{code}
\pause
\begin{code}
class Regex(pattern: String) {
  def unapply(s: String): Option[GroupsOf[@pattern@]]
}
\end{code}
\end{slide}

@@

\begin{slide}{What changed between 2016 and 2022?}
The Spark library uses a hack, implicits:
\begin{itemize}
  \item convoluted to use
  \item slow to compile
\end{itemize}

The regex library uses match types, a new language construct introduced in this thesis.
\end{slide}

@@

\begin{slide}{Generalized singleton types: a proposal}
Our proposal consists of 3 changes:
\begin{enumerate}
  \item new types for if-then-else, pattern matching, constructors, functions calls
  \item "precise mode" of type inference
  \item type evaluation, during subtyping
\end{enumerate}
\end{slide}

\begin{slide}{New types}
What's the type of | if (x == 0) "zero" else "one"|?
\begin{itemize}
  \pause\item |String| ~(Scala 2)
  \pause\item |"zero"|~{\footnotesize\texttt\textbar}~|"one"| ~(Scala 3)
  \pause\item |If[x.type == 0,|| "zero", "one"]| ~(proposed)
  \pause\item |{ if (x == 0) "zero" else "one" }| ~(proposed, syntactic sugar)
\end{itemize}
\pause
Similarly, we add new types for other constructs: |Match[]|, |New[]|, |Call[]|, |TypeTest[]|, and the corresponding syntactic sugar.
\end{slide}

@@

\begin{slide}{"Precise mode" of type inference: list concatenation example}
\begin{code}
dependent def concat(xs: List, ys: List) <: List =
  xs match
    case x :: xs => x :: concat(xs, ys)
    case Nil => ys
\end{code}
\begin{overprint}
\onslide<1-2>
\begin{code}
dependent val l1 = "A" :: Nil
dependent val l2 = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { "A" :: "B" :: Nil }
\end{code}

@@

More interesting for pattern matching: we "desugar" pattern matching expressions into if-then-else \& type tests.
\end{slide}
\begin{slide}{Type evaluation: type tests}
We evaluate type tests using subtyping and type disjointness:
\begin{itemize}
  \item |eval(@\tp{x.isInstanceOf}@[T]) = true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
  \item |eval(@\tp{x.isInstanceOf}@[T]) = false| \quad if |x.type| and |@\tp{T}@| are disjoint
  \item |eval(@\tp{x.isInstanceOf}@[T]) = @\tp{x.isInstanceOf}@[T]| \quad otherwise
\end{itemize}
\end{slide}

\begin{slide}{Type evaluation: pattern matching example}
\begin{overprint}
\onslide<1>
\begin{code}
dependent val foo(x: Any) =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<2>
\begin{code}
dependent val foo(x: Any): {
  x match
    case s: String => @\tp{s}@
    @\tp{case i: Int}@ @=>@ @\tp{i+1}@
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\onslide<3->
\begin{code}
dependent val foo(x: Any): {
  if (x.isInstanceOf[String]) x.asInstanceOf[String]
  @\tp{else if }@(@\tp{x.isInstanceOf}@[@\tp{Int}@]) @\tp{x.asInstanceOf}@[Int] @\tp{+ 1}@
  @\tp{else throw new}@ MatchError()
} =
  x match
    case s: String => s
    case i: Int => i+1
\end{code}
\end{overprint}
\bigskip
\pause\pause\pause

|foo(42): { 43 }|
\pause

|foo(readInt(): { (_: Int) + 1 })|
\end{slide}

\begin{slide}{Generalized singletons: recap}

We proposed a generalization of Scala's singleton types:
\begin{enumerate}
  \item lift a subset of Scala's language constructs to the type level
  \item add a "precise mode" of type inference
  \item evaluate types during subtyping
\end{enumerate}
\end{slide}

@@

\begin{slide}{Pattern matching on types}
\begin{code}
type Elem[X] = X match
  case String => Char
  case List[t] => Elem[t]
  case Any => X
\end{code}

\medskip

\begin{overprint}
\onslide<2-5>
\begin{code}
Elem[String] =:= Char
\end{code}

\onslide<3-5>
\begin{code}
Elem[String] =:= Char
Elem[Int] =:= Int
\end{code}

\onslide<4-5>
\begin{code}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[List[Int]] =:= Int
\end{code}

\onslide<5>
\begin{code}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[List[Int]] =:= Int
Elem[Seq[Int]] =:= Elem[Seq[Int]]
\end{code}

\onslide<6>
We evaluate type tests using subtyping and type disjointness:
\begin{itemize}
  \item |eval(@\tp{x.isInstanceOf}@[T]) = true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
  \item |eval(@\tp{x.isInstanceOf}@[T]) = false| \quad if |x.type| and |@\tp{T}@| are disjoint
  \item |eval(@\tp{x.isInstanceOf}@[T]) = @\tp{x.isInstanceOf}@[T]| \quad otherwise
\end{itemize}

\onslide<7>
Match types can be explained in terms of generalized singletons:

\medskip
\begin{code}
type Elem[X] = { (_: X) match
  case _: String => (_: Char)
  case _: List[t] @=>@ (_: Elem[t])
  case _: Any => (_: X)
}
\end{code}

\onslide<8>
The type/term correspondence dictates match type semantics:

\medskip
\begin{code}
def elem[X <: Singleton](x: X): Elem[X] = x match
  case x: String => x.charAt(0)
  case x: List[t] @=>@ elem(x.head)
  case x: Any => x
\end{code}

\onslide<9>
The type/term correspondence dictates match type semantics:

\medskip
\begin{code}
def elem[X <: Singleton](x: X): Elem[X] = x match
  case x: String => x.charAt(0)
  case x: List[t] => elem(x.head)
  case x: Any => x

elem[Any]("foo")
\end{code}
\vspace{-20pt}
\end{overprint}
\end{slide}

\begin{slide}{Formalization}
System FM is an extension of System \Fsub with
\begin{itemize}
  \item pattern matching
  \item opaque classes
  \item match types
\end{itemize}
\pause
Design goals:
\begin{itemize}
  \item explain the essence of match types
  \item give us confidence in our design
  \item be as simple as possible
\end{itemize}
\end{slide}

@@

\begin{tabular}{@{}llll@{}}
|class A; class B extends @\tp{A}@| &$\C = \{\A, \B\}$ & $\Psi = \{(\B, \A)\}$ & $\Xi = \varnothing$\\
|class A; class D| &$\C = \{\A, \D\}$ & $\Psi = \varnothing$  & $\Xi = \{(\A, \D)\}$\\
|class A; trait T| &$\C = \{\A, \T\}$ & $\Psi = \varnothing$  & $\Xi = \varnothing$\\
\end{tabular}
\end{slide}

@@

\begin{slide}{Empty types}
Our system does not like empty types.

@@

\begin{slide}{Match types: recap}
We presented match types, a new language construct for type-level programming:
\begin{itemize}
  \item implemented in the Scala 3 compiler
  \item formalized in an extension of System \Fsub
  \item already in active use
\end{itemize}
\end{slide}

@@

\begin{slide}{Type-safe regular expressions (simplified)}
\begin{code}
val rational = new Regex("(\\d+)\\.(\\d+)")
rational.unapply("3.1415"): Option[(String, String)]

@@

\end{document}
